Region-network hierarchical sparsity priors for\\ high-dimensional inference in brain imaging

### 2016-02-02 paper/paper.tex
sanitize figs
next figure sanitized
pass on methods

\begin{abstract}
Structured sparsity penalization has recently improved
statistical models applied to high-dimensional data in various domains.
As an extension to imaging neuroscience,
priors on network hierarchies of brain regions have been incorporated into
logistic-regression estimators for neural activity maps.
This remarries the perspectives of
functional segregation and functional integration
that are routinely divorced in neuroscientific research.
%
Region-network hierarchical priors for supervised classification
of 18 psychological tasks
from a reference dataset are
shown to outperform na\"ive $\ell_1$-norm and
trace-norm penalization, as well as neurobiologically
informed $\ell_1/\ell_2$-block-norm group sparsity.
Varying the relative importance of
region and network structure within the hierarchical tree prior
recovered complementary aspects of the neural activity patterns
with similar predictive accuracies.
%
In sum,
we demonstrate that priors of local and global neurobiological knowledge can enhance
out-of-sample performance and domain interpretability
by hierarchical tree sparsity.

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., ``long data")  to
the high-dimensional regime (i.e., ``wide data")
\cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This high-dimensional scenario challenges various statistical methods from
classical statistics.
For instance,
estimating generalized linear models without additional assumptions
yields an underdetermined system of equations.
%
Many such ill-posed estimation problems
have benefited from
\textit{sparsity} assumptions
\cite{buhlmann2011statistics, hastie2015statistical}.
They act as a 
regularizer and can be used for model selection.
Sparse supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Generally, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus shrinking and selecting variables guided by
expected data distributions
\cite{bach2012optimization}.
Such restrictions to complexity are an attractive plan of attack
for the $\textgreater$100,000 variables of brain maps.
Yet, what generally accepted neurobiological structure lends itself
to fruitful exploitation using structured sparsity priors?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions are responsible for distinct classes of computational 
processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of anatomy (i.e., histological preparation),
connectivity (i.e., axonal tracing),
and functional properties
(i.e., single-cell recordings) in animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local
%cyto- and chemoarchitectonic
infrastructure
and its unique global connectivity profile
together are thought to realize brain function.
%
In sum,
cortical brain modules versus connections between them
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation}.
Importantly,
probably no existing brain analysis method acknowledges that
both functional organizations are inextricably involved
in the realization of mental operations
\cite{tononi1998complexity, saygin2012}.

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single-cell recordings and microscopic examination
revealed, for instance, the
specialization in the occipital visual cortex into V1, V2, V3, V3A/B, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The whole-brain localization of
sensory, motor, and emotional functions to cortical areas
% in the living brain
was later enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptors
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical microscopical borders
\cite{behrens03, cbp2015review}.
High-thoughput approaches today enable
ultrahigh-resolution 3D models of brain anatomy
at near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.

% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015, stephan_dys}.
%
Invasive axonal tracing studies in monkeys were complemented
by diffusion MRI tractography in humans
as a now frequently employed method to
outline fiber bundles between brain regions
\cite{jbabdi2013long}.
Besides analyses of
electrophysiological oscillations
\cite{buzsaki2004neuronal}
and
graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
These revealed the important implication of
canonical brain networks across psychological tasks,
including the so-called
``default-mode network'' \cite{raichle2001pnas},
``salience network'' \cite{seeley2007dissociable},
and ``dorsal attention network'' \cite{corbettashul2008}. 
Characteristic changes in the configuration of
these macroscopical networks
were repeatedly observed to be induced
by the onset of given psychological tasks \cite{fransson2006}.
% Such task-induced mechanisms orchestrating supraordinate networks
% might be subserved by the right anterior insula \cite{sridh2008}
% and temporo-parietal junction \cite{bzdok2013tpj}.
%
As a common point of all these methods,
interpretation of findings naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in stark constrast to methods examining regional specialization.

% study
Building on these two major interpretational traditions in systems neuroscience,
the present study proposes to incorporate
established neurobiological structure underlying
functional segregation and integration
into supervised estimators
by hierarchical structured sparsity.
% The "true" relative importance of 
% local region compartments and global network compartments
% is typically unknown
% but
% probably varying in degree
% across diverse neuroscientific questions.
%
Learning techniques
exploiting structured sparsity 
have recently made much progress in various application domains
from processing auditory signals \cite{daudet2004sparse},
natural images \cite{harzallah2009combining} and
videos \cite{kang2015structured, kim2010sparse}
to
astrophysics \cite{vinci2014estimating},
genetics \cite{rapaport2008classification, kim2012tree},
and
conformational dynamics of protein complexes \cite{jenatton2009structured}.
%
This is extended by the present work that
introduced neuroscience-specific estimators capitalizing on
neurobiologically plausible region and network priors.
%
Based on the largest neuroimaging repository,
we demonstrated that domain-informed supervised models
gracefully tackle the curse of dimensionality,
yield more human-interpretable results,
and generalize better to new samples
than domain-na\"ive estimators.

\section{Methods}
Our main contribution is the domain-specific adaptation of sparse structured
penalties making it possible to jointly incorporate functional 
specialization and functional integration priors into statistical 
estimators.
We capitalize on hierarchical group lasso as introduced by Jennaton and 
others in \cite{jenatton2011structured} to create a set of convex penalty terms.
These can probe the interplay between regional specialization and 
inter-regional functional integration during defined psychological tasks.
%
\paragraph{Rationale.}
Three-dimensional voxel brain images as obtained by various neuroimaging
techniques are very high-dimensional, but also very highly structured.
While its explicit dimensionality, the number of brain voxels, which varies with
the resolution of the image is generally of the order of 100,000, the number
of samples available for analyses rarely exceeds hundreds or thousands.
This \(n << p\) scenario immediately implies underdetermination of any
linear model based on dot products with the voxel values.

Yet, there is abundant structure to be exploited, which can be
injected into the estimation as a domain knowledge prior. Here we are 
interested in incorporating knowledge about both the well-established 
modules of functional specialization and the spatiotemporal interactions
between these.
%
Developmentally, such large-scale networks emerge during late
fetal growth (Doria et al., 2010), before mental capacities mature
in childhood. 
%
In adults, nodes of a such a spatiotemporally cohesive network have more
similar functional profiles than nodes from different networks
(Anderson et al., 2013).
%
On the signal level, functional modules are localized structures implicating
neighboring voxels in an image. A variety of known functional modules has
been compiled into atlases of the brain. Different types of atlases rely on
different ways of segregating modules from another. For instance, anatomical
atlases rely on the variation of tissue properties across the brain, which
can be at the cellular level or based on higher-order structure.
%
Global functional networks can be discovered by statistical analysis of
functional brain imaging data, for example the BOLD contrast in fMRI. A
well-established method of estimating these networks is independent 
component analysis \cite{Hamalainen}. It robustly identifies the default
mode, saliency and attention networks.
%
Global functional networks as estimated by ICA are continuous brain maps
attributing weights to each voxel. Since the regions involved in these
networks are nevertheless largely disjoint spatially and roughly in
correspondence with module boundaries {\color{red} Danilo: Ok!}, 
it is possible to uniquely associate each module from a region atlas to one
of the extracted global functional networks.

This correspondence makes it possible to inject established knowledge about the 
networks and regions into a hierarchy: A brain contains a certain number of
networks, which in turn consist of the atlas modules as subregions.
We can use this hierarchy to create a structural prior of expected weight
maps for classifiers.

The hierarchical group lasso as introduced by \cite{jenatton2011} lends
itself perfectly to represent this structure. It is based on the group lasso
\cite{group_lasso} but permits groups to contain each other in a tree
structure. The first level of the tree are the network groups containing all
the voxels of the modules associated to them. Then each network contains as
as subgroups the modules associated with it, grouping the voxels of each of
these regions together.
As with the group lasso, it is possible to associate an individual penalty
to each group. In trading off the penalties on the network level against the
penalties of the region level, we can create different regimes of 
estimation: 
%
Setting a low penalty on the network groups makes it probable
that all of them are active in the estimated weight map. If we then select
higher penalties on region groups, selection of relevant region groups is
forced without the bias of the network maps.
%
Conversely, setting low penalties on the region maps makes it possible for
all voxels to be active. Selecting higher penalties on the networks then
leads to a selection of networks with all regions associated to it.
%
Taken together and varied smoothly, we can evaluate a tradeoff between the
relevance of modular regions and functionally integrated networks in 
statistical estimation of brain activity.

%% we need to inject domain knowledge into
%% statistical estimations to harness the curse of dimensionality.
%% two neurobiological design principles 
%% 
%% imposing parsimony
%% integrative processes
%% 
%%  This L1/L2 norm for group lasso has been extended to a more general setting to
%% 
%% designed groups
%%  the child nodes enter the set of relevant inputs only if its parent node does. 
%%  
%% should be able to estimate voxel level
%% while taking into account known supravoxel structure.
%% is instrumental in
%% Developmentally, such large-scale networks emerge during late
%% fetal growth (Doria et al., 2010), before cognitive capacities mature
%% in childhood. 
%% 
%% In adults, nodes of a same cohesive network have more
%% similar functional profiles than nodes from different networks
%% (Anderson et al., 2013).
%% 
%% data exhibit natural correlations between neighboring voxels forming clusters
%% 
%% representing some phenomenon with as few variables as possible
%% 
%% neurobiologically motivated restrictions to complexity circumvented
%% the curse of dimensionality
%% three-dimensional spatial arrangement that respects
%% the functional anatomy of the brain
%% not ignore the spatial configuration
%% 
%% incorporate rich prior knowledge
%% 
%% If meaningful structures exist,
%% we show that one can take advantage of such structures
%% 
%% Statistically,
%% l1 and l2 are local sparsity priors
%% -> resulting sparsity does yield structure
%% we want to priviledge representations with structure
%% 
%% 
%% => 
%%  a biologically and statistically desirable bias

@@

For a brain image \(w\in\R^p\) and a group \(g\), the vector 
\(w_g\in\R^{|g|}\) is defined as the restriction of \(w\) to the coordinates
 in \(g\). The structured penalty incorporating network
and region information can then be written as
\[\Omega(w) = \alpha\sum_{h\in\mathcal H}\eta_h\|w_h\|_2 + \beta\sum_{g\in\mathcal G}\eta_g\|w_g\|_2.\]
According to \cite{Zhou_or_who_was_that} we set \(\eta_g = 1/\sqrt{|g|}\) to
account for varying group size. The hierarchy-level-specific factors 
\(\alpha > 0\) and \(\beta > 0\) are used to trade-off region-weighted and
network-weighted models against each other.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/reg_net_prior.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Building blocks of the region-network tree.}
Depicts neurobiological priors introduced into the classification problem 
by hierarchical structured sparsity.
\textit{Left:} Continuous, partially overlapping brain network priors
(\textit{hot-colored}, taken from \cite{smith2009})
accommodate the functional integration
perspective of brain organization.
\textit{Right:} Discrete, non-overlapping brain region priors
(\textit{single-colored}, taken from \cite{crad12})
accommodate the functional segregation perspective.
\textit{Middle:} These two types of predefined voxel groups are incorporated
into hierarchical priors of parent networks with their
descending region nodes.
\textit{Top to bottom:} Four examplary region-network priors
are shown, including
the early cortex that processes
visual and sound information from the environment,
a well-known attentional circuit in the left brain hemisphere,
and
the cerebellum that is involved in motor behavior.
}
\label{fig_priors}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/sparsities.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Model performance across sparsity priors.}
Compares the performance of logistic regression estimators
with 4 different structured and unstructered sparsity terms
in classifying neural activity from 18 psychological tasks.
The class-wise precision and recall metrics
were obtained on the same test set.
%
Unstructured $\ell_1$-penalized logistic regression (\textit{upper left})
imposed a minimum of relevant brain voxels without
assuming special structure.
Structured $\ell_1/\ell_2$-block norm with additional $\ell_1$ term
(\textit{lower left})
imposed region compartments, but na\"ive to network structure.
Structured trace-norm penalization (\textit{lower right})
imposed low-rank structure
with sparsity of network patterns, but na\"ive to region structure.
Structured $\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors (\textit{upper right})
exhibited the best out-of-sample performance.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\label{fig_sparsities}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/dataratio.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Na\"ive versus informed sparse model selection
across training set sizes.}
Ordinary $\ell_1$-penalized logistic regression
(\textit{upper row})
is compared
to hierarchical-tree-penalized logistic regression
(\textit{lower row})
with increasing fraction
of the available training data (\textit{left to right columns}).
For one example (i.e., ``View tools'') from 18 psychological tasks,
unthresholded axial maps of model weights
are shown for comparison against
the sample average of that class
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
The out-of-sample accuracies for predicting all 18 psychological tasks
is given in percent.
%
In the data-scarce scenario,
typical for brain imaging,
hierarchical tree sparsity achieves much
better support recovery with the biggest difference
in model performance.
%
In the data-rich scenario,
neurobiologically informed logistic regression
profits more from the available information quantities than
neurobiologically naive logistic regression.
}
\label{fig_dataratio}
\end{figure}

\paragraph{Sample complexity of na\"ive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
were quantitatively compared (Figure \ref{fig_dataratio}).
Region-network priors should bias model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-na\"ive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP task data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 neural activity maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).
%
The stratified and shuffled training data were submitted
to a nested cross-validation scheme
for model selection and model assessment.
In the inner CV layer, the logistic regression estimators
have been trained in a one-versus-rest design that
distinguishes each class from
the respective 17 other classes
(number of maximal iterations=$100$, tolerance=$0.001$).
In the outer CV layer, grid search
selected among candidates for the respective $\lambda$ parameter
by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
Importantly, the thus selected sparse logistic regression classifier was
evaluated on an identical test set in all settings.

>>>>

\begin{abstract}
Structured sparsity penalization has recently improved
statistical models applied to high-dimensional data in various domains.
As an extension to imaging neuroscience,
the present work incorporates
priors on network hierarchies of brain regions into
logistic-regression estimators to distinguish neural activity.
This remarries the perspectives of
functional segregation and functional integration
that are routinely divorced in neuroscientific research.
%
Region-network hierarchical priors for supervised classification
of 18 psychological tasks
from a reference dataset are
shown to outperform naive $\ell_1$-norm and
trace-norm penalization, as well as neurobiologically
informed $\ell_1/\ell_2$-block-norm group sparsity.
Weighing the relative importance of
region and network structure within the hierarchical tree penalty
recovered complementary aspects of the neural activity patterns.
%
It is thus demonstrated
that priors of local and global neurobiological knowledge
can enhance
generalization performance, sample complexity, and domain interpretability
by hierarchical tree sparsity.

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., ``long data")  to
the high-dimensional regime (i.e., ``wide data").
% the high-dimensional regime (i.e., ``wide data") \cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This high-dimensional scenario challenges various statistical methods from
classical statistics.
For instance,
estimating generalized linear models without additional assumptions
yields an underdetermined system of equations.
%
Many such ill-posed estimation problems
have benefited from
\textit{sparsity} assumptions
\cite{hastie2015statistical}.
They act as a 
regularizer and can be used for model selection.
Sparse supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Generally, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus shrinking and selecting variables guided by
expected data distributions
\cite{bach2012optimization}.
Such restrictions to statistical complexity
are an attractive plan of attack
for the $\textgreater$100,000 variables per brain map.
Yet, what generally accepted neurobiological structure lends itself
to fruitful exploitation using structured sparsity priors?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions are responsible for distinct classes of computational 
processes
\cite{kanwisher2010functional}.
Conversely, functional integration emphasizes that brain function
is enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
Local
infrastructure
and its unique global connectivity profile are thought to go hand-in-hand
to realize brain function.
%
However,
probably no existing brain analysis method acknowledges that
both functional organizations are inextricably involved
in the realization of mental operations.

Functional specialization has been
explored and interpreted based on many different research methods.
%
For instance,
single-cell recordings and microscopic examination
revealed the
specialization in the occipital visual cortex into V1, V2, V3, V3A/B, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.
% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015}.
%
Besides analyses of
electrophysiological oscillations
% \cite{buzsaki2004neuronal}
and
graph-theoretical properties,
% graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity correlation \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
%
As a common point of all these methods,
interpretation of findings naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in stark constrast to methods examining regional specialization.

% % specialization & integration
% Concepts on human brain organization have long been torn
% between the two extremes
% \textit{functional specialization} and \textit{functional integration}.
% Functional specialization emphasizes that microscopically distinguishable
% brain regions are responsible for distinct classes of computational 
% processes
% \cite{kanwisher2010functional}.
% Functional integration, in turn, emphasizes that brain function
% is enabled by complex connections between these
% distinct brain regions \cite{sporns14nn}.
% %
% These notions were predominantly derived from
% invasive examination of anatomy (i.e., histological preparation),
% connectivity (i.e., axonal tracing),
% and functional properties
% (i.e., single-cell recordings) in animals.
% Regarding functional segregation into specialized regions,
% early histological investigations into the microscopic heterogeneity of
% the human cerebral cortex have resulted
% in several detailed anatomical maps
% \cite{brodmann1909vergleichende}.
% Regarding axonal connections,
% each such cortical area has been observed
% to possess a unique set of incoming and outgoing connections
% \cite{passingham2002, young93monkey, scannell95cat}.
% %
% Both
% local
% %cyto- and chemoarchitectonic
% infrastructure
% and its unique global connectivity profile
% together are thought to realize brain function.
% %
% In sum,
% cortical brain modules versus connections between them
% reflect 
% functional specialization versus functional integration
% \cite{friston2002beyond, mesulam_sensation}.
% Importantly,
% probably no existing brain analysis method acknowledges that
% both functional organizations are inextricably involved
% in the realization of mental operations
% \cite{tononi1998complexity, saygin2012}.
% 
% 
% 
% % SPECIAL
% Functional specialization has been
% explored and interpreted based on many different research methods.
% %
% Single-cell recordings and microscopic examination
% revealed, for instance, the
% specialization in the occipital visual cortex into V1, V2, V3, V3A/B, and V4
% \cite{hubel1962receptive, zeki1978functional}.
% Tissue lesion of the mid-fusiform gyrus of the visual system,
% in turn,
% was frequently reported to impair
% recognition of others' identity from faces
% \cite{iaria2008contrib}.
% The whole-brain localization of
% sensory, motor, and emotional functions to cortical areas
% % in the living brain
% was later enabled by
% non-invasive brain imaging with
% functional magnetic resonance imaging (fMRI) and
% positron emission tomography (PET)
% \cite{fristen1997imaging}.
% Further,
% radioactive mapping of neurotransmitter receptors
% rendered accessible yet another
% local characteristic of neuronal populations
% \cite{zilles2009receptor}.
% % allude to our many-regions layer in our methods
% In the computational era,
% automatic clustering methods are increasingly employed to
% regionally differentiate the cerebral cortex,
% which can partly be more fine-grained than
% classical microscopical borders
% \cite{behrens03, cbp2015review}.
% High-thoughput approaches today enable
% ultrahigh-resolution 3D models of brain anatomy
% at near-cellular scale
% \cite{amunts2013bigbrain}.
% %
% As a crucial common point,
% all these methodological approaches
% yield neuroscientific findings
% that are naturally interpreted according to
% non-overlapping, discrete region compartments
% as the basic architecture of brain organization.
% 
% 
% 
% % INTEGRAL
% It is more recent
% that the main interpretational focus has shifted
% from circumscribed regions to network stratifications
% in systems neuroscience \cite{yuste2015, stephan_dys}.
% %
% Invasive axonal tracing studies in monkeys were complemented
% by diffusion MRI tractography in humans
% as a now frequently employed method to
% outline fiber bundles between brain regions
% \cite{jbabdi2013long}.
% Besides analyses of
% electrophysiological oscillations
% \cite{buzsaki2004neuronal}
% and
% graph-theoretical properties \cite{bullmore2009complex},
% studies of
% functional connectivity \cite{buckner2013opportunities} and
% independent component analysis (ICA) \cite{beckmann2005}
% became the workhorses of network discovery
% in neuroimaging.
% These revealed the important implication of
% canonical brain networks across psychological tasks,
% including the so-called
% ``default-mode network'' \cite{raichle2001pnas},
% ``salience network'' \cite{seeley2007dissociable},
% and ``dorsal attention network'' \cite{corbettashul2008}. 
% Characteristic changes in the configuration of
% these macroscopical networks
% were repeatedly observed to be induced
% by the onset of given psychological tasks \cite{fransson2006}.
% % Such task-induced mechanisms orchestrating supraordinate networks
% % might be subserved by the right anterior insula \cite{sridh2008}
% % and temporo-parietal junction \cite{bzdok2013tpj}.
% %
% As a common point of all these methods,
% interpretation of findings naturally embraces
% cross-regional integration by
% overlapping network compartments
% as the basic architecture of brain organization,
% in stark constrast to methods examining regional specialization.

% study
Building on these two major interpretational traditions in systems neuroscience,
the present study proposes to incorporate
established neurobiological structure underlying
functional segregation and integration
into supervised estimators
by hierarchical structured sparsity.
% The "true" relative importance of 
% local region compartments and global network compartments
% is typically unknown
% but
% probably varying in degree
% across diverse neuroscientific questions.
%
Learning techniques
exploiting structured sparsity 
have recently made much progress in various application domains
from processing auditory signals \cite{daudet2004sparse},
natural images \cite{harzallah2009combining} and
videos \cite{kang2015structured}
% videos \cite{kang2015structured, kim2010sparse}
to
astrophysics \cite{vinci2014estimating},
genetics \cite{kim2012tree},
% genetics \cite{rapaport2008classification, kim2012tree},
and
conformational dynamics of protein complexes \cite{jenatton2009structured}.
%
The recently suggested hierarchical tree penalties for imaging neuroscience
\cite{jenatton2011multi} will be extended
by introducing neurobiologically plausible region and network priors
to design neuroscience-specific estimators.
%
Based on the largest neuroimaging repository,
we demonstrated that domain-informed supervised models
gracefully tackle the curse of dimensionality,
yield more human-interpretable results,
and generalize better to new samples
than domain-naive estimators.

\section{Methods}
The main contribution is the domain adaptation of hierarchical structured
tree penalties to jointly incorporate region 
specialization and network integration priors into supervised 
estimators.
We capitalize on hierarchical group lasso as recently introduced
\cite{jenatton2011structured} to create a set of convex penalty terms.
These allow acknowledging the interplay between local specialization and 
global connectivity during defined psychological tasks.
%
\paragraph{Rationale.}
3D brain maps obtained by neuroimaging
techniques are high-dimensional but also highly structured.
While its explicit dimensionality, the number of brain voxels,
typically exceeds 100,000 variables, the number
of samples rarely exceeds few hundreds.
This \(n << p\) scenario directly implies underdetermination of any
linear model based on dot products with the voxel values.

Yet, the abundant structure can be exploited by
injecting domain-knowledge priors into the statistical estimation process.
Here we are 
interested in well-established neurobiological knowledge about
brain regions of functional specialization and spatiotemporal interactions
between these.
%
These large-scale brain networks emerge in human individuals around birth
\cite{doria2010}, before mental capacities mature in childhood. 
In adults,
the sets of regions that compose the spatiotemporally cohesive network have more
similar functional profiles than nodes from different networks
\cite{anderson2013}.
As a well-established method for network estimation
\cite{beckmann2005},
ICA yields continuous brain maps with
weights for each voxel. The region nodes of ICA
networks are spatially disjoint and in
correspondence with boundaries of brain atlases.
Consequently,
each region from a brain atlas can be uniquely associate to one
the extracted ICA networks.
%
The network regions are spatially compact structures of
neighboring voxels in a neural activity map.
The brain's grey matter can be segregated into functional regions to
compile brain atlases
using microscopical or algorithmic investigations
\cite{cbp2015review}.
%
This brain-network correspondence
allows constructing a hierarchy of global ICA networks with their
assigned local regions.
This network-region tree is used as a structural prior
of expected weight distributions
for supervised learning algorithms.

The hierarchical group lasso introduced by \cite{jenatton2011multi} lends
itself perfectly to represent this tree structure.
It extends the group lasso
\cite{yuan2006model} to permit variable groups to contain each other
in a nested tree structure.
The first hierarchical level of that tree are
the network groups that contain all
the voxels of the modules associated to them.
Each network node, in turn, descends into a set of
brain regions associated with spatially neighboring voxel groups.
%
As an important consequence,
a child node enters the set of relevant voxel variables only
if its parent node has been selected.
Conversely,
if a parent node is deselected,
also the voxel variables of all child nodes are deselected.
%
Moreover,
the coefficient of the region and network groups can
be weighed individually.
Trading off the voxel penalties of the network level against the
voxel penalties of the region level,
we can design different estimation regimes.
%
Setting a low penalty on the network groups makes it probable
that all of them are active in the estimated weight map. If we then select
higher penalties on region groups, selection of relevant region groups is
forced without the bias of the network maps.
Conversely, setting low penalties on the region maps makes it possible for
all voxels to be active. Selecting higher penalties on the networks then
leads to a selection of networks with all regions associated to it.
%
This region-network tradeoff allows inspecting the relative contributions
of the region and network levels
of the hierarchical prior.

%%  a biologically and statistically desirable bias

@@

For a brain image \(w\in\R^p\) and a group \(g\), the vector 
\(w_g\in\R^{|g|}\) is defined as the restriction of \(w\) to the coordinates
 in \(g\). The structured penalty incorporating network
and region information can then be written as
\[\Omega(w) = \alpha\sum_{h\in\mathcal H}\eta_h\|w_h\|_2 + \beta\sum_{g\in\mathcal G}\eta_g\|w_g\|_2.\]
According to \cite{yuan2006model} we set \(\eta_g = 1/\sqrt{|g|}\) to
account for varying group size. The hierarchy-level-specific factors 
\(\alpha > 0\) and \(\beta > 0\) are used to trade-off region-weighted and
network-weighted models against each other.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/reg_net_prior_colin.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Building blocks of the region-network tree.}
Depicts neurobiological priors introduced into the classification problem 
by hierarchical structured sparsity.
\textit{Left:} Continuous, partially overlapping brain network priors
(\textit{hot-colored}, taken from \cite{smith2009})
accommodate the functional integration
perspective of brain organization.
\textit{Right:} Discrete, non-overlapping brain region priors
(\textit{single-colored}, taken from \cite{crad12})
accommodate the functional segregation perspective.
\textit{Middle:} These two types of predefined voxel groups are incorporated
into hierarchical priors of parent networks with their
descending region nodes.
\textit{Top to bottom:} Four examplary region-network priors
are shown, including
the early cortex that processes
visual and sound information from the environment,
a well-known attentional circuit in the left brain hemisphere,
and
the cerebellum that is involved in motor behavior.
}
\label{fig_priors}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/sparsities.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Model performance across sparsity priors.}
Compares the performance of logistic regression estimators
with 4 different structured and unstructered sparsity terms
in classifying neural activity from 18 psychological tasks.
The class-wise precision and recall metrics
were obtained on the same test set.
%
Unstructured $\ell_1$-penalized logistic regression (\textit{upper left})
imposed a minimum of relevant brain voxels without
assuming special structure.
Structured $\ell_1/\ell_2$-block norm with additional $\ell_1$ term
(\textit{lower left})
imposed region compartments, but naive to network structure.
Structured trace-norm penalization (\textit{lower right})
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
Structured $\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors (\textit{upper right})
exhibited the best out-of-sample performance.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\label{fig_sparsities}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/dataratio_trans.pdf}
\end{centering}
\vspace{-0.6cm}
\caption{\textbf{Naive versus informed sparse model selection
across training set sizes.}
Ordinary $\ell_1$-penalized logistic regression
(\textit{upper row})
is compared
to hierarchical-tree-penalized logistic regression
(\textit{lower row})
with increasing fraction
of the available training data (\textit{left to right columns}).
For one example (i.e., ``View tools'') from 18 psychological tasks,
unthresholded axial maps of model weights
are shown for comparison against
the sample average of that class
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
The out-of-sample accuracies for predicting all 18 psychological tasks
is given in percent.
%
In the data-scarce scenario,
typical for brain imaging,
hierarchical tree sparsity achieves much
better support recovery with the biggest difference
in model performance.
%
In the data-rich scenario,
neurobiologically informed logistic regression
profits more from the available information quantities than
neurobiologically naive logistic regression.
}
\label{fig_dataratio}
\end{figure}

\paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
were quantitatively compared (Figure \ref{fig_dataratio}).
Region-network priors should bias model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP task data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 neural activity maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).
%
The stratified and shuffled training data were submitted
to a nested cross-validation scheme
for model selection and model assessment.
In the inner CV layer, the logistic regression estimators
have been trained in a one-versus-rest design that
distinguishes each class from
the respective 17 other classes
(number of maximal iterations=$100$, tolerance=$0.001$).
In the outer CV layer, grid search
selected among candidates for the respective $\lambda$ parameter
by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
Importantly, the thus selected sparse logistic regression classifier was
evaluated on an identical test set in all settings.
